\ppse{} - A Peer-to-Peer Support for Multiplayer Games

### 2008-12-31 trunk/textos/p2pse-sbgames2008/ppsg.tex
@
funfando no linux

% \author{JEMS Manuscript Number 48534}
% \contactinfo{48534@JEMS}
%% Author and Affiliation (multiple authors). Use: and between authors
\author{Felipe J. Vilanova, Carlos Eduardo B. Bezerra , Marcos R. Crippa, Fábio R. Cecin, Cláudio F. R. Geyer\\
          Universidade Federal do Rio Grande do Sul\\
          Bento Gonçalves, 9500\\
          Porto Alegre, RS, Brazil\\
          +55 (51) 3316 0000
}
\contactinfo{\{fjvilanova, carlos.bezerra, mrcrippa, fcecin, geyer\}inf.ufrgs.br}

@@

  The successful commercial MMOGs, at the moment, are implemented using some variant of the client-server model. This model offers simple design, good security and fast detection and treatment of cheating. However, it lacks scalability. The cost of maintaining the server becomes excessive with the increased number of customers. One approach to deal with this situation is distributing the MMOG simulation. The changes in the virtual world of the game would be processed by the machines of the clients, without interference from the server. This distribution can be achieved using the peer-to-peer model. In this paper, we describe a network engine, called \ppse{}, which supports partially decentralized MMOGs, ensuring scalability and flexibility, while guaranteeing the basic properties of a client-server model, such as security and consistency. We first examine in more details the existent distributed models. After that, we discuss about the \ppse{} project and its characteristics. Finally we present a simulation of the architecture in comparison with the traditional client-server model.

@@

Multiplayer games are a very popular game genre, due to their highly interactive nature. Among those games, a class that is growing in popularity is MMOGs (Massively Multiplayer Online Games). MMOGs are real-time multiplayer games played through the Internet, where a large number of players (usually thousands) play within a persistent-state virtual world. Successful examples include EverQuest \cite{everquest} and World of Warcraft \cite{worldofwarcraft}.

The most common network model used in MMOGs is client-server. In this model, the client only sends its data to the server-side (that can operate on a single machine, a cluster or a grid) and receives frequent updates from the server. The server process all the data from the clients, and broadcasts all the results that occur on the virtual world back to them. Some advantages of this model are: being simple in design; cheating can be efficiently detect and stopped; retaining control over access to the game; and being a predictable model.

@@

The client-server paradigm is undoubtedly the most currently used in the implementation of Internet real-time multiplayer games. Commercial examples of these games include Doom \cite{doom}, Quake \cite{quake} and Counter-Strike \cite{counterstrike}, all them 3D real time action games. In any client-server game, massively multiplayer or not, the players interact with the environment and with other players through a client program. Each client has a network connection only with the server. The server is responsible for receiving the information from each client and passing updates to the other clients. There are several different ways the communication protocol between the client and the server can be implemented: some of them provide more security, while others provide efficiency.

Regardless of the technique used, the server is responsible for maintaining the game state up-to-date. Since the game is a real time simulation, it is necessary for the state of the game to be frequently recalculated. In general, the server is configured to update its state in a fixed and relatively small frequency, and to periodically send update messages to all clients. Due to that, the server must have great processing power (CPU and, indirectly, memory) and enough bandwidth (network).

A distributed game can use the centralized simulation to accommodate thousands of simultaneous clients. In practice, however, it can be difficult to have in a single machine all the processing power required to perform the simulation in real time. One solution to this problem, commonly used in commercial MMOGs such as EverQuest and PlanetSide \cite{planetside}, is to divide the virtual world and simulate each division in an individual machine as, for example, a particular computer in a cluster of computers. But even using the clusters to mitigate the problem of the server processing cost, the problem of network consumption still remains. The MMOG server, a single machine or a cluster, will need an Internet connection with low latency and very high bandwidth.

@@

The Communications Architecture for Massive Multiplayer Games, proposed by \cite{fiedler2002cam}, was the first architecture to address the scalability problem by proposing a solution based on the publisher-subscriber paradigm. In this paradigm, the virtual world is divided into smaller pieces, usually called "cells", and each participant can choose to sign (or participate) in only a few cells. Thus, each subscriber of a cell only needs to exchange update messages with subscribers of the same cell. These architectures also deal with other fundamental problems, such as the problem of responsiveness to commands generated by each participant.

@@

Guild Wars, as far as we can tell, follows the client-server model. Our proposal, which is illustrated in Figure \ref{fig:instanced}, is to use the instanced game model with a hybrid network architecture model. Since the interactions on the social space (chatting, trading) don't require high consistency, and there is a necessity for validation of who can play (game accounts, passwords), the server-side will be responsible for coordinating this part of the game. So even a server-side with modest processing capacity and network bandwidth could manage the social space if game quality is scaled down accordingly, for instance by sending less frequent updates to clients. The game sessions would be processed only on the client-side machines, avoiding unnecessary processing and communication cost on the server-side. The clients would form groups, which would operate within a peer-to-peer dynamics.

@@

The first layer in the \ppse{} architecture is responsible for forming the mesh of point-to-point connections. The clients' connections with the servers and with other clients are established and managed in this layer. The basic network functions for connection establishment, such as socket creation, connection and packet sending and receiving, are given by a low-level API provided by the ZIG library \cite{zige} which basically implements a protocol that resembles SCTP \cite{stewart2000sct} over UDP. The term ``connection'', when used in the context of the \ppse{} library, refers to this ZIG-layer level protocol connection, which, again, is implemented over UDP, adding some features, such as reliability, timeout detection and utilization of only one single socket in every host to handle all communications.

The Layer 0 server is responsible for managing connections with all clients. Naturally, it maintains a list of all clients that are connected to it along with a logical communication endpoint object for each client. When started, the server spawns a socialization space called the social space object. The clients that are not in an action space are bound to the social space - or to the ``limbo'', when in transit from one space to the other. All communication between clients that are in the social space is intermediated by the server.

The server can create new empty groups in order to instantiate new action spaces, or destroy such existing groups and their corresponding spaces. In the action spaces all the clients communicate directly, which requires them to already have the ability to send UDP packets to one another. The entry of a client into a group is managed by the server in the following manner:

\begin{enumerate}
\item The server sends a synchronization message to the new member and to all the other members informing them to which clients they must be connected in order to enter (or stay) in the group. The server sends a list with all the members of the group;
\item All members try then to establish connections to every other member in the recently received member list, or just keep the connection, if it has already been established (the ZIG library supports simultaneous bidirectional connections through the same socket);
\item When a client is entering a group, but still has not finished the process, it belongs to the ``limbo'';
\item As soon as the new member establishes connections to the other clients, the server excludes it from the limbo and inserts it in the member list of the group;
\item When a client is removed from the group, all connections to the other clients are interrupted and it returns to the social space;
\item If a client, for any reason, loses its connection to the server, it is removed from the group;
\item When a client enters or leaves a group, the server sends to the others  an up-to-date member list.
\end{enumerate}

@@

The ns-2 simulator \cite{mccanne:nsn} was used to create the simulated environment and perform the simulation itself. The servers and clients - both traditional and \ppse{} - were programmed extending the Application base class of the ns-2 API. The \ppse{} clients connect to the \ppse{} server, sending packets to it at a fixed rate while in the social space. The server, in turn, sends to each client in the social space an update containing information about all other clients in the same space. When a player leaves the social space and joins a peer-to-peer group of an action space, it stops exchanging packets with the server, due to its direct communication with the other players. The non-\ppse{} server and clients behave simply as follows: each client sends a packet to the server at a given fixed rate, while the server broadcasts an update to all clients, also at a fixed rate. Although this approach does not take into consideration possible optimizations on the server, such as interest management \cite{morse1996iml}, this is not relevant, since the same optimizations could be used in the social space of the \ppse{} server, resulting in a similar bandwidth usage reduction percentage, when comparing the two server models.

@@

We proposed a different approach to MMOGs, that uses P2P groups and transfers the simulation processing to them. In order to provide security and scalability, the game model was restrict to the instanced game model. The existence of a server-side  guarantees that, if deem necessary, the server can act as final arbiter. The performed simulations demonstrate that, even in a pessimistic scenario, the reduction of the average bandwidth usage by the game server can be significant. The download bandwidth utilization was reduced by one half, while the upload bandwidth utilization was the most benefited one, decreasing by 75\%, allowing to reduce the maintenance cost of such kind of server.

We are currently integrating the \ppse{} architecture with our game prototype, \hoverkill{}. This game is a client-server capture-the-flag tank game, and its network system is being replaced by our architecture. The final goal is to be able to test the game in client-server and peer-to-peer modes, in order to base our conclusions in real results rather than simulations.

@@

This work was supported by the Funding for Studies and Projects (FINEP), through the \ppse{} project (\ppse{}-5849-1), and by the National Research Council (CNPq).

>>>>

% \author{JEMS Manuscript Number 48534}
% \contactinfo{48534@JEMS}
%% Author and Affiliation (multiple authors). Use: and between authors
\author{Felipe J. Vilanova, Carlos Eduardo B. Bezerra , Marcos R. Crippa, Fábio R. Cecin, Cláudio F. R. Geyer\\
          Universidade Federal do Rio Grande do Sul\\
          Bento Gonçalves, 9500\\
          Porto Alegre, RS, Brazil\\
          +55 (51) 3316 0000
}
\contactinfo{\{fjvilanova, carlos.bezerra, mrcrippa, fcecin, geyer\}@inf.ufrgs.br}

@@

  The successful commercial MMOGs, at the moment, are implemented using some variant of the client-server model. This model offers simple design, good security and fast detection and treatment of cheating. However, it lacks scalability. The cost of maintaining the server becomes excessive with the increased number of customers. One approach to deal with this situation is distributing the MMOG simulation. The changes in the virtual world of the game would be processed by the machines of the clients, without interference from the server. This distribution can be achieved using the peer-to-peer model. In this paper, we describe a network engine, called \ppse{}, which supports partially decentralized MMOGs, ensuring scalability and flexibility, while guaranteeing the basic properties of a client-server model, such as security and consistency. We first examine more closely the existing distributed models. After that, we discuss the \ppse{} project and its characteristics. Finally we present a simulation of the architecture in comparison with the traditional client-server model.

@@

Multiplayer is a very popular game genre, due to its highly interactive nature. Among those games, MMOGs (Massively Multiplayer Online Games) are becoming increasingly popular. MMOGs are real-time multiplayer games played through the Internet, where a large number of players (usually thousands) play within a persistent-state virtual world. Successful examples include EverQuest \cite{everquest} and World of Warcraft \cite{worldofwarcraft}.

The most common network model used in MMOGs is client-server. In this model, the client only sends its data to the server (which can operate on a single machine, a cluster or a grid) and receives frequent updates from the server. The server process all the data from the clients, and broadcasts all the results that occur on the virtual world back to them. Some advantages of this model are: being simple in design; cheating can be efficiently detected and stopped; retaining control over access to the game; and being a predictable model.

@@

The client-server paradigm is undoubtedly the most used one in the implementation of Internet real-time multiplayer games. Commercial examples of these games include Doom \cite{doom}, Quake \cite{quake} and Counter-Strike \cite{counterstrike}, all of them 3D real time action games. In any client-server game, massively multiplayer or not, the players interact with the environment and with other players through a client program. Each client has a network connection only with the server. The server is responsible for receiving the information from each client and passing updates to the other clients. There are several different ways the communication protocol between the client and the server can be implemented: some of them provide more security, while others provide efficiency.

Regardless of the technique used, the server is responsible for keeping the game state up-to-date. Since the game is a real time simulation, it is necessary for the state of the game to be frequently recalculated. In general, the server is configured to update its state in a fixed and relatively small frequency, and to periodically send update messages to all clients. Due to that, the server must have great processing power (CPU and, indirectly, memory) and enough bandwidth (network).

A distributed game can use the centralized simulation to accommodate thousands of simultaneous clients. In practice, however, it can be difficult to have all the processing power required to perform the simulation in real time in a single machine. One solution to this problem, commonly used in commercial MMOGs such as EverQuest and PlanetSide \cite{planetside}, is to divide the virtual world and simulate each piece of it on an individual machine such as a particular computer in a cluster. But even using the clusters to mitigate the problem of the server processing cost, the problem of network consumption still remains. The MMOG server, a single machine or a cluster, will need an Internet connection with low latency and very high bandwidth.

@@

The Communications Architecture for Massive Multiplayer Games, proposed by \cite{fiedler2002cam}, was the first architecture to address the scalability problem by proposing a solution based on the publisher-subscriber paradigm. In this paradigm, the virtual world is divided into smaller pieces, usually called ``cells'', and each participant can choose to sign (or participate) in only a few cells. Thus, each subscriber of a cell only needs to exchange update messages with subscribers of the same cell. These architectures also deal with other fundamental problems, such as the lack of responsiveness to commands generated by each participant problem.

@@

Guild Wars follows the client-server model. Our proposal, which is illustrated in Figure \ref{fig:instanced}, is to use the instanced game model with a hybrid network architecture model. Since the interactions on the social space (chatting, trading) don't require high consistency, and there is a necessity for validation of who can play (game accounts, passwords), the server-side will be responsible for coordinating this part of the game. So even a server-side with modest processing capacity and network bandwidth could manage the social space if game quality is scaled down accordingly, for instance by sending less frequent updates to clients. The game sessions would be processed only on the client-side machines, avoiding unnecessary processing and communication cost on the server-side. The clients would form groups, which would operate within a peer-to-peer dynamics.

@@

The first layer in the \ppse{} architecture is responsible for forming the mesh of point-to-point connections. The clients' connections to the servers and to other clients are established and managed in this layer. The basic network functions for connection establishment, such as socket creation, connection and packet sending and receiving, are given by a low-level API provided by the ZIG library \cite{zige} which basically implements a protocol that resembles SCTP \cite{stewart2000sct} over UDP. The term ``connection'', when used in the context of the \ppse{} library, refers to this ZIG-layer level protocol connection, which, again, is implemented over UDP, adding some features, such as reliability, timeout detection and utilization of only one single socket in every host to handle all communications.

The Layer 0 server is responsible for managing connections with all clients. Naturally, it maintains a list of all clients that are connected to it along with a logical communication endpoint object for each client. When started, the server spawns a socialization space called the social space object. The clients that are not in an action space are bound to the social space - or to the ``limbo'', when in transit from one space to the other. All communication between clients that are in the social space passes through the server.

The server can create new empty groups in order to instantiate new action spaces, or destroy existing groups and their corresponding spaces. In the action spaces all the clients communicate directly, which requires them to already have the ability to send UDP packets to one another. The entry of a client into a group is managed by the server in the following manner:

\begin{enumerate}
\item The server sends a synchronization message to the new member and to all the other members informing them to which clients they must be connected in order to enter (or stay) in the group. The server sends a list with all the members of the group;
\item All members try then to establish connections to every other member in the recently received member list, or just keep the connection, if it has already been established (the ZIG library supports simultaneous bidirectional connections through the same socket);
\item When a client is entering a group, but still has not finished the process, it belongs to the ``limbo'';
\item As soon as the new member establishes connections to the other clients, the server excludes it from the limbo and inserts it in the member list of the group;
\item When a client is removed from the group, all of its connections to the other clients are interrupted and it returns to the social space;
\item If a client, for any reason, loses its connection to the server, it is removed from the group;
\item When a client enters or leaves a group, the server sends an up-to-date member list to the others.
\end{enumerate}

@@

The ns-2 simulator \cite{mccanne:nsn} was used to create the simulated environment and perform the simulation itself. The servers and clients - both traditional and \ppse{} - were programmed extending the Application base class of the ns-2 API. The \ppse{} clients connect to the \ppse{} server, sending packets to it at a fixed rate while in the social space. The server, in turn, sends to each client in the social space an update containing information about all other clients in the same space. When a player leaves the social space and joins a peer-to-peer group of an action space, it stops exchanging packets with the server, due to its direct communication with the other players. The non-\ppse{} server and clients behave simply as follows: each client sends a packet to the server at a given fixed rate, while the server broadcasts an update to all clients, also at a fixed rate. Although this approach does not take into consideration possible optimizations on the server, such as interest management \cite{morse1996iml}, this is not relevant, since the same optimizations could be used in the social space of the \ppse{} server, resulting in a similar percentage of bandwidth usage reduction, when comparing the two server models.

@@

We proposed a different approach to MMOGs, that uses P2P groups and transfers the simulation processing to them. In order to provide security and scalability, the game model was restricted to the instanced game model. The existence of a server-side  guarantees that, if deemed necessary, the server can act as a final arbiter. The performed simulations demonstrate that, even in a pessimistic scenario, the reduction of the average bandwidth usage by the game server can be significant. The download bandwidth utilization was reduced by one half, while the upload bandwidth utilization was the most benefited one, decreasing by 75\%, allowing to reduce the maintenance cost of such kind of server.

We are currently integrating the \ppse{} architecture with our game prototype, \hoverkill{}. This game is a client-server capture-the-flag tank game, and its network system is being replaced by our architecture. The final goal is to be able to perform tests under a real environment, in order to base our conclusions in real results rather than simulations.

@@

This work was supported by the Funding for Studies and Projects (FINEP), through the \ppse{} project (\ppse{}-5849-1), and by the National Research Council (CNPq). We also would like to thank Andrius Batalauskas (Hochschule Bremen) for the final review of the text.
